The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-12 article.tex
Intro gets better
Introduction

The computational tools a contemporary phylogeneticist has at his/her disposal
goes beyond the wildest imagination of those living three decades ago. 
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in 1981 \cite{felsenstein1981}. The first Bayesian tools 
appeared in 1996 \cite{rannala1996}, allowing
for unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
MrBayes \cite{mrbayes} and RevBayes \cite{revbayes}. 
Each of these use efficient tree sampling 
routines to create a posterior probability distribution of the to-be-inferred 
phylogenetic tree. But next to the computational speed, they also allow
to incorporate known or hypothesized structure of the phylogenetic 
tree-to-be-inferred through the model priors.

In a Bayesian analysis, the model priors are explicitly specified.
Those model priors can be grouped in three groups: (1) site model, which
governs the nucleotide subsititution model, (2) clock model, specifying
the rate of mutation per lineage in time, and (3) tree prior, embodying
the speciation model behind branching events (speciation) 
and branch termination (extinction).

The effect of choosing a (potentially wrong) prior will have its effect on
the posterior. Recently, for example, \cite{moller2018} show that the effect
of choosing a tree prior biases the estimation of the molecular clock rate, 
for DNA sequences of 100-1000 base pairs.

This bias has its implications in the research and policy regarding rapidly 
evolving viruses. [NOTE: I would say that same study concludes 
that it does not matter for DNA alignments of 15K, which is the DNA 
alignment length that I use].

All of the currently implemented tree priors assume that speciation is 
instantaneous, where we know that speciation can be a gradual process.
The protracted birth-death (PBD) model, an extension of 
the (constant-rate) birth-death (BD) model, incorporates this idea.
In the PBD model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, or finish its speciation to become a good species.

Although we know speciation takes time, all of our theoretical 
phylogenetic modeling tools ignore this fact. One reason may be to ease
computation, enabling researchers to work on bigger phylogenies (with
more power statistical power) instead. Additionally, speciation being
protracted is only playing a role in the present. As seen from the present,
all older species have finished speciation, as we -by definition- recognize
these species as such.

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of an underestimated amount of species. 
But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the most recent common 
ancestor (MRCA) of a young species duo is estimated 
too close to the present, as in the BD model a species can be 
ultra-young (as speciation is assumed to be instantaneous), 
where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

From a theoretical point of view, there is another reason why protracted
speciation has not yet become a member of our toolbox: there is no framework
where it fits in: a framework assumes either an analysis at the species or
subspecies level. When picking a tool that does its analysis at the
species level, there is no protocol how to select which incipient species
gets to represent the full species it is still a member of. When picking
a tool at the sub-species level, the incipient state disappears as a
sub-species lineage, and the process degrades to an ordinary birth-death 
process.

This exploratory research's goal is to produce a 
useful and balanced data set, that quantifies the inference error 
we make from choosing the incorrect prior. In this context, 
we measure the error made on PBD phylogenies, when using a BD prior.
In brief, we simulate protracted phylogenies using the protracted birth-death process,
from those phylogenies sample a species tree. From that species tree, 
we simulate a DNA sequence alignment. Then we let BEAST infer a phylogeny 
from these alignments using the standard birth-death prior. 
We compare the posterior phylogenies to the species tree using the nLTT statistic.

>>>>

The computational tools a contemporary phylogeneticist has at his/her disposal
goes beyond the wildest imagination of those living three decades ago. 
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in 1981 (\cite{felsenstein1981}). The first Bayesian tools 
appeared in 1996 (\cite{rannala1996}), allowing
for unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are BEAST (\cite{beast})
and its successor BEAST2 (\cite{beast2}),  
MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
Each of these use efficient tree sampling 
routines to create a posterior probability distribution of the to-be-inferred 
phylogenetic tree. But next to the computational speed, they also allow
to incorporate known or hypothesized structure of the phylogenetic 
tree-to-be-inferred through the model priors.

In a Bayesian analysis, the model priors are explicitly specified.
Those model priors can be grouped in three groups: (1) site model, which
governs the nucleotide subsititution model, (2) clock model, specifying
the rate of mutation per lineage in time, and (3) tree prior, embodying
the speciation model behind branching events (speciation) 
and branch termination (extinction).
The effect of choosing a (potentially wrong) prior will have its effect on
the posterior. For example, recently, it was shown that the effect
of choosing a tree prior biases the estimation of the molecular clock rate, 
for DNA sequences of 100-1000 base pairs (\cite{moller2018}).

At the present time, for sexually producing animals, all we have at our
disposal are tree priors that assume that speciation
is instantaneous, where we know that speciation is a gradual process.
When assuming big populations sizes (thus the effect of sampling to 
be small), the (constant-rate) birth-death (BD) model is a commonly 
used tree prior, which does not incorporate the temporal aspect of speciation.
The protracted birth-death (PBD) model, an extension of 
the BD model, does incorporate the idea that speciation takes time.
In this model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, finish its speciation to become a good species, or give
rise to new incipient species.

The effect of using the (incorrect) BD tree prior for a PBD process is unknown.
A potential problem in species conservation is that the number of 
species is underestimated (see \cite{fennessy2016} for a clear example). 
Additionally, protracted speciation may be one 
explanation in the observed decline of speciation rates in 
time (\cite{etienne2012}). Also, a BD model places the most recent common 
ancestor (MRCA) of a young species duo closer to the present, as the
BD model allows for a speciation event being recognized immediately,
where the PBD model acknowledges that this will take time.

There are multiple possiblities why the PBD model is relatively unexplored.
Biologically, the PBD model is predicted to have an effect strongest in
the present (as earlier speciation events are nearly always recognized),
so in research that investigates (mostly) older species, a BD model would suffice. 
Computationally, the BD model is simpler, thus more convenient, model. 
Methodological, there is no computational tool where 
the PBD model fits in: every contemporary framework assumes 
either an analysis at the species or subspecies level. In the PBD
model, incipient species are the cause there is no such thing 
as a 'true' species tree, as incipient species may give rise to paraphylies.
It may be possible to establish a protocol to sample a species tree from an 
incipient species tree, but this has not yet been explored [NOTE: except by Simonet].

This research's goal is to explore the effect of using an overly simplistic
BD prior on a PBD simulated reality. We provide a data set, 
that quantifies the inference error made in general, and explores
the effect of the way species trees are sampled from an incipient species tree.  
In brief, we simulate protracted phylogenies using the PBD process,
from which we sample a species tree. From that species tree, 
we simulate a DNA sequence alignment. Then, BEAST2 uses these alignments
to infer a posterior of phylogenies, using a BD prior. The difference
between the (BD) posterior phylogenies and (PBD) species tree is quantified.

% Methods
